Countering language attrition with PanLex and the Web of Data

### 2014-03-10 panlex-rdf-paper/2012/MLOD_PanLex/paper_v5.tex
updated fig 2
- switched to figure version without fake data
- used bigger foaf:homepage URL with ellipsis

\begin{figure}
\includegraphics[width=0.46\textwidth]{images/pdf/ontology_new.pdf}
% Query for ontology_new.pdf
%SELECT * WHERE {
%  ?den plx:denotationMeaning ?mea .
%  ?den plx:denotationExpression ?expr .
%  ?expr rdfs:label ?exprlbl .
%  ?mea plx:meaningDefinition ?def .
%  ?mea plx:approver ?app.
%  ?def plx:languageVariety <http://ld.panlex.org/plx/languageVariety/187> .
%  ?def rdfs:label ?dtx .
%  ?den plx:denotationExpression <http://ld.panlex.org/plx/expression/355621> .
%  ?app dc:title ?dctitle .
%  ?app plx:quality ?plxquality .
%  ?app dc:publisher ?dcpublisher .
%  ?app dc:date ?dcdate .
%  ?app rdfs:label ?rdfslabel .
%  ?app foaf:homepage ?foafhomepage .
%  ?app dc:creator ?dccreator .
%  ?app plx:registrationDate ?regdate .
%  ?app plx:license <http://ld.panlex.org/plx/license/mi> .
%  <http://ld.panlex.org/plx/license/mi> rdfs:label ?liclbl .
%  OPTIONAL { ?app dbpedia-owl:isbn ?dboisbn . }
%  OPTIONAL { ?den plx:wordClass ?wordclass . }
%}
\caption{Example of the PanLex RDF vocabulary showing one meaning of the expression `between' and the corresponding source and definition}
\label{fig:vocabulary}
\end{figure}

@@

\section{Linking}
\label{sec:linking}
The SML view in (\autoref{fig:ex:sparqlify-ml}) establishes the interlinking of languages in PanLex with Lexvo.
Here we outline the interlinking with DBpedia, where we were interested in creating \emph{valid} and \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38}}, which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and removal of punctuation characters.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.

@@

\section{Dataset Benefits and Usage Scenarios}
\label{sec:usage}
There are general benefits of using Semantic Web technologies, such as the
potential for simplified data integration due to RDF and vocabulary reuse, the
possibility of enriching data based on interlinking, drawing advantage from
reasoning and the exploration of the data through the use of generic Semantic
Web tools.
Moreover, some applications, like the TeraDict translation lookup
service\footnote{\url{http://panlex.org/teradict/?lg=eng}}, can now be realized
using SPARQL queries and so easily integrated in other applications.
Due to space considerations, we refer the reader to the PanLex Linked Data
landing page\footnote{\url{http://ld.panlex.org}}, where a
collection of SPARQL queries is maintained.
Also, since PanLex covers a niche of providing linguistic data for
non-mainstream languages, investigation of its fitness for use in cross
language information retrieval, as well as annotation projects, like DBpedia
Spotlight\footnote{\url{http://spotlight.dbpedia.org}}, seems worthwhile.

@@

\section{Related Work}
\label{sec:related}
PanLex is a project whose editors integrate information discovered from many lexical resources.
The extraction of information from linguistic sources, and techniques for automatically inferring translations,
are relevant work discussed in \cite{panlex_probtrans}.
%A similar approach of linking expressions to meanings is followed by the
% \emph{Global Wordnet Association}.
An important related initiative is the Global Wordnet Association (GWA)\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing wordnets and, among other goals, uniformly representing wordnets
of different languages and establishing a universal index of meaning.
Wordnets are usually focused on the
definition of \emph{synsets} and relations between them in a single language; GWA is helping to transform these single-language synonym silos into a virtual multilingual translation resource. PanLex is approximating that, in a different way, by integrating data from numerous wordnets along with translingual sources into a single graph.
%As already mentioned in PanLex there can be several meaning entities
% representing the same unique sense.
In the Semantic Web context, several standard or quasi-standard vocabularies and ontologies have been developed with the rise of the Linguistic LOD movement.
Examples include the
\emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010} for modeling
lexicon and machine-readable dictionaries, \emph{POWLA} for modeling linguistic corpora\cite{powla2012} and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

>>>>

\begin{figure}
\includegraphics[width=0.46\textwidth]{images/pdf/ontology_new3.pdf}
% Query for ontology_new.pdf
%SELECT * WHERE {
%  ?den plx:denotationMeaning ?mea .
%  ?den plx:denotationExpression ?expr .
%  ?expr rdfs:label ?exprlbl .
%  ?mea plx:meaningDefinition ?def .
%  ?mea plx:approver ?app.
%  ?def plx:languageVariety <http://ld.panlex.org/plx/languageVariety/187> .
%  ?def rdfs:label ?dtx .
%  ?den plx:denotationExpression <http://ld.panlex.org/plx/expression/355621> .
%  ?app dc:title ?dctitle .
%  ?app plx:quality ?plxquality .
%  ?app dc:publisher ?dcpublisher .
%  ?app dc:date ?dcdate .
%  ?app rdfs:label ?rdfslabel .
%  ?app foaf:homepage ?foafhomepage .
%  ?app dc:creator ?dccreator .
%  ?app plx:registrationDate ?regdate .
%  ?app plx:license <http://ld.panlex.org/plx/license/mi> .
%  <http://ld.panlex.org/plx/license/mi> rdfs:label ?liclbl .
%  OPTIONAL { ?app dbpedia-owl:isbn ?dboisbn . }
%  OPTIONAL { ?den plx:wordClass ?wordclass . }
%}
\caption{Example of the PanLex RDF vocabulary showing one meaning of the expression `between' and the corresponding source and definition}
\label{fig:vocabulary}
\end{figure}

@@

\section{Linking}
\label{sec:linking}
The SML view in (\autoref{fig:ex:sparqlify-ml}) establishes the interlinking of languages in PanLex with Lexvo~\cite{deMeloWeikum2008c}.
Here we outline the interlinking with DBpedia~\cite{dbpedia-swj}, where we were interested in creating \emph{valid} and \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38}}, which map (non-localized) DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels by applying Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and removal of punctuation characters.
Each DBpedia resource was then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
\autoref{fig:plx-dbp-link-counts} summarizes the number of links obtained.

@@

\section{Dataset Benefits and Usage Scenarios}
\label{sec:usage}
There are general benefits of using Semantic Web technologies, such as the
potential for simplified data integration due to RDF and vocabulary reuse, the
possibility of enriching data based on interlinking, drawing advantage from
reasoning and the exploration of the data through the use of generic Semantic
Web tools.
Moreover, some applications, like the TeraDict translation lookup
service\footnote{\url{http://panlex.org/teradict/?lg=eng}}, can now be realized
using SPARQL queries and so easily integrated in other applications.
Due to space considerations, we refer the reader to the PanLex Linked Data
landing page\footnote{\url{http://ld.panlex.org}}, where a
collection of SPARQL queries is maintained.
Also, since PanLex covers a niche of providing linguistic data for
non-mainstream languages, investigation of its fitness for use in cross
language information retrieval, as well as annotation projects, like DBpedia
%Spotlight\footnote{\url{http://spotlight.dbpedia.org}},
Spotlight~\cite{isem2011mendesetal}
 seems worthwhile.

@@

\section{Related Work}
\label{sec:related}
PanLex is a project whose editors integrate information discovered from many lexical resources.
The extraction of information from linguistic sources, and techniques for automatically inferring translations,
are relevant work discussed in~\cite{panlex_probtrans}.
%A similar approach of linking expressions to meanings is followed by the
% \emph{Global Wordnet Association}.
An important related initiative is the Global Wordnet Association (GWA)\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing wordnets and, among other goals, uniformly representing wordnets
of different languages and establishing a universal index of meaning.
Wordnets are usually focused on the
definition of \emph{synsets} and relations between them in a single language; GWA is helping to transform these single-language synonym silos into a virtual multilingual translation resource. PanLex is approximating that, in a different way, by integrating data from numerous wordnets along with translingual sources into a single graph.
%As already mentioned in PanLex there can be several meaning entities
% representing the same unique sense.
In the Semantic Web context, several standard or quasi-standard vocabularies and ontologies have been developed with the rise of the Linguistic LOD movement.
Examples include the
\emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010} for modeling
lexicon and machine-readable dictionaries, \emph{POWLA} for modeling linguistic corpora~\cite{powla2012} and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.
